---
title: "Lab 2 - Data Structures in R"
subtitle: "Environmental Data Analytics"
author: "John Fay & Luana Lima"
---

## Objective
Discuss and navigate different data types in R
 - Vectors
 - Matrices
 - Arrays
 - Data frames

---

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: <https://www.statmethods.net/input/datatypes.html>.

-   **Vectors** 1 dimensional structure that contains elements of the same type.

-   **Matrices** 2 dimensional structure that contains elements of the same type.

-   **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

-   **Lists** Ordered collection of elements that can have different modes.

-   **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

---

## Vectors
### Creating Vectors
Add comments below that specify what kind of vectors are created. (First answer is provided.)

```{r Creating vectors}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # Vector type: Numeric
vector1
class(vector1)

vector2 <- c("one","two","three") # Vector type: String
vector2
class(vector2)

vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) # Vector type: Boolean
vector3
class(vector3)
```

### Vector Operations
Add comments below to indicate what each operation does.(First answer is provided.)

```{r Vector operatons}
# Returns the 3rd item in the vector
vector1[3]

#Returns the maximum value of the first vector
max(vector1)

#Create another vector of the values 10 thru 15
vector4 <- c(10:15)

#Combines the two vectors
c(vector1, vector4)

#Multiplies the vectors together linear algebra style
vector1 * vector4

#Multiplies vector1 by 10
vector1 * 10

#Makes vector 2 into all uppercase
toupper(vector2) 

#Returns the opposite
!vector3

```

### Exercise 1a: Vector
Find a 5-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create a vector representing the high temperature on each of the five days.
<https://www.wunderground.com/forecast/us/nc/durham/KNCDURHA138>

```{r Vector Exercise}

#Assign to the variable 'temps_F' a vector of high temperatures in F
temps_F <- c(97, 97, 96, 83, 84)
temps_F

#What is the average of these temperatures 
average_temp_F <- mean(temps_F)
average_temp_F

#Compute the difference between each temperature and the average calculated above
difference_in_temps <- c(temps_F - average_temp)
difference_in_temps
temps_F-mean(temps_F)

#Subtract 32 and multiply by 5/9ths to create a vector in Celsius
temps_C <- c((temps_F - 32) / 1.8)
temps_C

```

## Matrices
###Creating Matrices
Complete the comments below:
```{r Creating Matrices}
# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # 
matrix1
#matrix1 has 5 rows and 4 columns

matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #
matrix2
#matrix2 has 5 rows and 4 columns

matrix3 <- matrix(
  data = 1:20, 
  nrow = 5, 
  ncol = 4, 
  byrow = TRUE, # return after comma continues the line
  dimnames = list(
    c("uno", "dos", "tres", "quatro", "cinco"),
    c("un", "deux", "trois", "quatre")
    )
  ) 
matrix3
```

### Matrix Operations
Add comments in the code chunk below to indicate type of vectors created and what each operation does.
```{r Matrix Operations}

matrix1[4, 1] #

matrix1[4,  ] #

matrix1[ , 3] #

matrix1[c(12, 14)] #

matrix1[c(12:14)] #

matrix1[2:4, 1:3] #

#Create a vector of 6 values
values = c(1, 26, 24, 68, 3, 44)
#Create a vector of row names
rnames <- c("R1", "R2") 
#Create a vector of column names
cnames <- c("C1", "C2", "C3") 

matrix4 <- matrix(
  data = values, 
  nrow = 2, 
  ncol = 3, 
  byrow = TRUE,
  dimnames = list(
    rnames, 
    cnames)
  ) # 

matrix4 #

matrix4['R2','C1']
```

### Exercise 1b: Matrices
Compute the mean of column "C2" in matrix4. (Should be 14.5)
```{r Matrix Exercise}
mean(matrix4[,3])
```



## Lists
### Creating Lists
```{r Creating Lists}
# Lists ---- 
# Creating Lists
list1 <- list(
  name = "Maria", 
  mynumbers = vector1, 
  mymatrix = matrix1, 
  age = 5.3)
list1

```


### List Operations
```{r List Operations}
# List Operations

#
list1

#
list1$age

#
names(list1)

#
list1[1:2]

#
list1$name = "Janelle"

#
list1[[4]] = 44
```

## Data Frames
### Creating dataframes
```{r Creating data frames}
# Data Frames ----
#Create three vectors of equal length
d <- c(1, 2, 3, 4) # What type of vector?
e <- c("red", "white", "red", NA) # What type of vector?
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector?

#Combine the vectors into a dataframe
dataframe1 <- data.frame(d,e,f) 
View(dataframe1)

#Set the column names & View
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # MUST comment out "View(dataframe1)" BEFORE knitting.

#Add a new column to the dataframe
dataframe1$month <- c('Jan','Feb', 'Mar', 'Apr')

View(dataframe1)# MUST comment out "View(dataframe1)" BEFORE knitting.

```


### Dataframe operations
```{r Dataframe Operations}
#Extract a column from the dataframe f
dataframe1$Color

#Extract a row
dataframe1[2,]

#Extract several rows
dataframe1[1:2,] 

#Extract a column
dataframe1[,1]

#Extract a column/columns from their name
out1 <- dataframe1['Color']; class(out1)
out1

out2 <- dataframe1$Color; class(out2)
out2

out3 <- dataframe1[c("ID","Passed")]; class(out3)
out3

```