---
title: "M3 - Lab 3 - Part 2 - Exploring a New Environmental Dataset"
subtitle: "Environmental Data Exploration"
author: "John Fay & Luana Lima"
output: pdf_document
---

## About this lab

In the lesson you explored a USGS streamflow dataset. Now run the **same
data-detective workflow** on a completely different environmental dataset —
**EPA daily PM2.5 (fine particulate matter) for North Carolina** — on your
own. Same moves: **set up → understand → inspect → diagnose → summarize → three
plots → write it up.** You are diagnosing, not cleaning.

---

## Stage 0 — Set up the session

- Load `tidyverse`, `lubridate`, and `here`.
- Import `EPAair_PM25_NC2025_raw.csv` with a project-relative path.
- Parse the `Date` column into a real date.

```{r setup, message = FALSE}
# Load packages


# Confirm the project root with here()


# Import the raw PM2.5 file (project-relative path via here())
pm25 <- read.csv(
  #---- ENTER THE FILE PATH HERE, using here() ----
)

# Parse the date (you learned date parsing in the lesson; these dates are clean,
# so it's a single line). First LOOK at a raw value: head(pm25$Date)
# EPA daily files are usually MM/DD/YYYY -> use mdy(). (day-first? use dmy())
pm25$Date <- mdy(pm25$Date)
```

---

## Stage 1 — Understand the source

Briefly: where do these data come from, what does one row represent, and in what
units is PM2.5 reported?

> Answer:

---

## Stage 2 — Inspect the structure

```{r inspect}
# glimpse the data; check dim() and names()


```

What stands out about the size, the columns, and the `Date` column?

> Answer:

---

## Stage 3 — Check types & identify the categorical variables

Which columns are categorical (grouping) variables rather than measurements?

```{r categoricals}
# Look at the unique values of the likely categorical columns
# (e.g., Source, UNITS) and count the sites (Site.ID)


# If you'll group or plot by site, make Site.ID a factor explicitly:
# pm25$Site.ID <- as.factor(pm25$Site.ID)
```

> Answer (which columns are categorical?):

---

## Stage 4 — Diagnose: sites, missingness, and coverage

```{r diagnose}
# How many monitoring sites are in the dataset?


# How many missing values per column? (colSums + is.na)


# What is the date range? (works once Date is parsed)


# Any exact duplicate rows?


```

> Answer:

---

## Stage 5 — Descriptive summaries

- How many observations are present?
- What is the mean (and median) daily PM2.5 concentration?
- In what units?

```{r summaries}
# number of observations


# mean and median daily PM2.5 (na.rm = TRUE), and a full summary()


# units


```

> Answer:

---

## Stage 6 — Three diagnostic plots

One categorical, one distribution, one relationship. Detective tools only —
customization is M5.

```{r plots, fig.height = 3, fig.width = 4, warning = FALSE}
# (1) CATEGORICAL — number of samples from each data Source (bar chart)


# (2) DISTRIBUTION — daily mean PM2.5 concentration (histogram)


# (3) RELATIONSHIP — concentration over time (Date on x; a time series)


```

What does each plot tell you?

> Answer:

---

## Stage 7 — Write a short data-exploration summary

> - **One key characteristic:**
> - **One data-quality issue:**
> - **One interesting pattern:**
> - **One question for further analysis:**
